---
title: "Assignment_4"
output: 
  html_document:
    toc: TRUE
    number_sections: TRUE
    theme: paper
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE, warning = FALSE}
library(gapminder)
library(tidyverse)
```

# Univariate Option 1

## Make a tibble with one row per year, and columns for life expectancy for two or more countries.

```{r}
gapminder %>%
  select(year, lifeExp, country) %>%
  filter(country == "Cambodia" | country == "Vietnam" | country == "Thailand") %>%
  pivot_wider(names_from = country,
              values_from = lifeExp) %>%
  DT::datatable()
```

## Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

```{R}
gapminder %>%
  select(year, lifeExp, country) %>%
  filter(country == "Cambodia" | country == "Vietnam" | country == "Thailand") %>%
  pivot_wider(names_from = country,
              values_from = lifeExp) %>%
  ggplot(aes(x = Cambodia, y = Vietnam)) +
    geom_point() +
    xlab(label = "Average Life Expectancy in Cambodia") +
    ylab(label = "Average Life Expectancy in Vietnam") +
    theme_bw()
```

The two are directly proportional except for two outlier years during which Cambodia had a much lower life expectancy than did Vietnam.

## Re-lengthen the data.

```{r}
gapminder %>%
  select(year, lifeExp, country) %>%
  filter(country == "Cambodia" | country == "Vietnam" | country == "Thailand") %>%
  pivot_wider(names_from = country,
              values_from = lifeExp) %>%
  pivot_longer(cols = c(Cambodia, Vietnam, Thailand),
               names_to = "country",
               values_to = "lifeExp") %>%
  DT::datatable()
```

The table is returned to its original length.

# Multivariate Option 1

## Make a tibble with one row per year, and columns for life expectancy and GDP per capita (or two other numeric variables) for two or more countries.

```{r}
gapminder %>%
  select(year, country, lifeExp, gdpPercap) %>%
  filter(country == "Brazil" | country == "Colombia") %>%
  pivot_wider(names_from = c(country),
              names_sep = "_",
              values_from = c(lifeExp, gdpPercap)) %>%
  DT::datatable()
  
```

## Re-lengthen the data.

```{r}
gapminder %>%
  select(year, country, lifeExp, gdpPercap) %>%
  filter(country == "Brazil" | country == "Colombia") %>%
  pivot_wider(names_from = c(country),
              names_sep = "_",
              values_from = c(lifeExp, gdpPercap)) %>%
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_") %>%
  DT::datatable()
```

The table is returned to it's original length.

# Table Joins

```{r}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```

## For each guest in the guestlist (guest tibble), add a column for email address, which can be found in the email tibble.

```{r}
email %>%
  separate_rows(guest, sep = ", ") %>%
  rename(name = guest) %>%
  right_join(guest, by = "name") %>%
  DT::datatable()
  
```

## Who do we have emails for, yet are not on the guestlist?

```{r}

email %>%
  separate_rows(guest, sep = ", ") %>%
  rename(name = guest) %>%
  anti_join(guest) %>%
  knitr::kable()

```

We have emails for Turner Jones, Albert Marshall, and Vivian Marshall, but they are not on the guest list.

## Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).

```{r}
email %>%
  separate_rows(guest, sep = ", ") %>%
  rename(name = guest) %>%
  full_join(guest, by = "name") %>%
  DT::datatable()
```

This table includes everyone we have emails for as well as everyone on the original guestlist.
